---
output: rmdformats::material
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning =  FALSE, message = FALSE, skimr_include_summary = FALSE)
```


```{r}
devtools::install_github('bbc/bbplot')

packages <-
  c(
    "here",
    "tidyverse",
    "skimr",
    "lubridate",
    "assertthat",
    "sjlabelled",
    "bbplot",
    "naniar"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```



# Data cleaning

## Cases

The main dataset contains the dates of arrest and disposition (trial or court appearance) during the period in which the program operated. The file also contains an indicator of whether the arrestee was referred to the intervention program for that arrest (i.e. whether they were treated), whether the person was re-arrested while awaiting trial, the number of prior arrests at the time of program entry, and the arrest location.

We will start by loading and inspecting the data:

```{r}
cases <-
  read.csv(
    here(
      "data",
      "raw",
      "case.csv"
    )
  )

skim(cases)
```

The dataset contains `r cases %>% n_distinct` unique observations over `r cases %>% nrow` rows. The unique identifier in this dataset is the `caseid`, containing `r cases %>% select(caseid) %>% n_distinct` unique values. `person_id`, on the other hand, contains `r cases %>% select(person_id) %>% n_distinct` unique values and is a foreign key that can be used to link it to other datasets.

To prepare the data for analysis, we will change the formats of each variable. Right now, they are all stored as either text or numbers. However, we can see that there are date in the data, and that the variables `treat` and `re_arrest` are dummies. Dummy variables are better handled by R is stored in logical format. Finally, addresses are not standardized: they vary in their usage of upper and lower cases. We will standardize them by making them all upper case.

```{r}
cases <-
  cases %>%
  mutate(
    # Change dates to date format
    across(
      ends_with("date"),
      ~ as_date(.)
    ),
    # Change dummies to logical format
    across(
      c(treat, re_arrest),
      ~ as.logical(.)
    ),
    # Make text all upper case
    address = str_to_upper(address)
  )
```

This data will be saved in R format for easier handling in the future.

```{r}
write_rds(
  cases,
  here(
    "data",
    "clean",
    "case.rds"
  )
)
```


## Demographic characteristics

The dataset called `demo.csv` contains demographic information about arrestees, including some who were not included in the program evaluation.

```{r}
demo <-
  read.csv(
    here(
      "data",
      "raw",
      "demo.csv"
    )
  )
```

The dataset contains `r demo %>% n_distinct` unique observations over `r demo %>% nrow` rows, meaning there are full duplicates in the dataset. The unique identifier in this dataset is the `person_id`, containing `r demo %>% select(person_id) %>% n_distinct` unique values. The first thing to do before using this dataset to make sure we don't have multiple observations for a single individual, as this could mean that these individuals will receive a higher weight in analysis.

```{r}
# Remove duplicate rows
demo <-
  demo %>%
  unique
```

We can now explore the dataset to understand how the information is organized.

```{r}
skim(demo)
```

The only numeric variable in this dataset is the ID. All other variables are stored as characters. However, we know that `bdate` is an individual's birth date, and should therefore be in date format. Additionally, race and gender both have only a few unique values, which means they should probably be handled as categorical variables.

```{r}
demo <-
  demo %>%
  mutate(
    # Transform categorical variables into factors
    across(
      c(race, gender),
      ~ as_factor(.)
    ),
    # Turn birth date into a date
    bdate = as_date(bdate)
  )

# Check the data again
skim(demo)
```

There are three categories in race (BLACK, WHITE, and ASIAN) and four categories in gender (M, F, Male and Female). We will standardize these categories to handle them correctly.

```{r}
demo <-
  demo %>%
  mutate(
    # Standardize different codes for same gender
    gender =
      gender %>%
      fct_recode(
        "male" = "M",
        "female" = "F"
      ),
    # Make all factors proper case
    across(
      c(race, gender),
      ~ str_to_title(.) %>%
        as_factor
    )
  )
```

Finally, let's save the data for future use.

```{r}
demo %>%
  write_rds(
    here(
      "data",
      "clean",
      "demo.rds"
    )
  )
```

## Prior arrests

This data set reflects the arrests among individuals in `case.csv` that occurred before the program implementation, between 2008-2011.

```{r}
prior_arrests <-
  read.csv(
    here(
      "data",
      "raw",
      "prior_arrests.csv"
    )
  )
```

There are `r prior_arrests %>% select(person_id) %>% n_distinct` unique person IDs and `r prior_arrests %>% unique %>% nrow` unique observations, meaning there are duplicate entries. This could be because a person was arrested more than once on the same day or because the same arrest is recorded twice in the data. We know from the data description the it is possible for the same individual to be arrested multiple times in the same day.

```{r}
skim(prior_arrests)
``` 

The arrest date is currently stored as text, which makes it harder to process it.  We will change its format and then save the data:

```{r}
prior_arrests <-
  prior_arrests %>%
  mutate(
    arrest_date = as_date(arrest_date)
  )
```

```{r}
prior_arrests %>%
  write_rds(
    here(
      "data",
      "clean",
      "prior_arrests.rds"
    )
  )
```

## Grades

The last data table we received includes 9th and 10th grade course grades for a subset of individuals in `case.csv`.

```{r}
grades <-
  read.csv(
    here(
      "data",
      "raw",
      "grades.csv"
    )
  )
```

The dataset contains `r grades %>% unique %>% nrow` unique observations over `r grades %>% nrow` r ows. It also contains data for `r grades %>% select(person_id) %>% n_distinct` unique individuals.

```{r}
skim(grades)
```

Note that the data is currently stored at the level of the individual, although grades are measured by grade, term and subject. To facilitate the handling of this data, we will reshape the dataset to the individual-grade-term-subject level.

```{r}
grades <-
  grades %>%
  pivot_longer(
    cols = gr9_fall_math:gr10_spring_hist,
    names_pattern = "gr(.*)_(.*)_(.*)",
    names_to = c("grade", "term", "subject"),
    values_to = "score"
  )
```

Finally, we will turn the grades into factors, since they represent categories, and remove entries for subjects that the individual was not attending on a given term.

```{r}
grades <-
  grades %>%
  filter(score != "") %>%
  mutate(
    across(
      c(grade, term, subject),
      ~ . %>% str_to_sentence %>% as.factor
    )
  )
```

Let's now save this data.

```{r}
grades %>%
  write_rds(
    here(
      "data",
      "clean",
      "grades.rds"
    )
  )
```

# Construction

Construct the dataset that will be used for analysis.

```{r}
cases <-
  read_rds(
    here(
      "data",
      "clean",
      "case.rds"
    )
  )

demo <-
  read_rds(
    here(
      "data",
      "clean",
      "demo.rds"
    )
  )

prior_arrests <-
  read_rds(
    here(
      "data",
      "clean",
      "prior_arrests.rds"
    )
  )

grades <-
  read_rds(
    here(
      "data",
      "clean",
      "grades.rds"
    )
  )
```

## Combine case and demographic data

We will combine the demographic dataset with the main dataset so demographic variables can be included in the analysis. We will use the variable `person_id` to link the information on the individuals present in the case dataset to the same individual on the demografic dataset.

```{r}
constructed <-
  cases %>%
  left_join(
    demo,
    by = "person_id"
  )
```

To confirm that the operation was successful, we need to check if it created any missing values in the constructed dataset. If there are no missings in the newly added variables, then all observations were successfully combined.

```{r}
constructed %>%
  select(race, gender, bdate) %>%
  skim()
```

## Restrict sample to individuals arrested in Chicago

While the program was mostly rolled out to defendants in Chicago, the State’s Attorney’s Office also ran a pilot serving a small number of individuals arrested in other parts of Cook County. For the purpose of this analysis, we will restrict the data to only individuals who were arrested in Chicago. To do this, we assume that the `address` variable indicates the address where the arrest occurred.

Right now, the city and the street address are stored in the same column, separated by a comma. To be able to filter observations based on the city where the arrest occurred, we will first need to separate the two.

```{r}
constructed <-
  constructed %>%
  # Separate city from street address
  separate(
    address, 
    sep = ", ",
    into = c("st_address", "city")
  ) %>%
  # Make city name proper case
  mutate(
    city = str_to_title(city)
  )
```

Let's check the results through a frequency table:

```{r}
table(constructed$city)
```

Now that we know the operation was successful, we can restrict our sample of interest to arrests that took place in Chicago.

```{r}
constructed <-
  constructed %>%
  filter(city == "Chicago")
```

## Defendant’s age at the time of arrest

We have information on the birth date for each defendant, as well as the date they were arrested, which allows us to calculate the age at the time of arrest

```{r}
constructed <-
  constructed %>%
  mutate(
    age = (bdate %--% arrest_date) / years(1)
  )
```

## GPA

The State’s Attorney is interested in pursuing a partnership with the Chicago Public Schools to investigate the relationship between high school achievement and criminal justice outcomes in early adulthood. To that end, the State’s Attorney’s Office has requested 9th and 10th grade course grade data from defendants between the ages of 18 and 24. We will calculate the GPA for each individual in each grade combining all terms and subjects in each grade. When constructing GPA, please use a 4 point scale, where: A=4, B=3, C=2, D=1, and F=0.

```{r}
gpa <-
  grades %>%
  # Convert grade to numeric score
  mutate(
    score = case_when(
      score == "F" ~ 0,
      score == "D" ~ 1,
      score == "C" ~ 2,
      score == "B" ~ 3,
      score == "A" ~ 4
    )
  ) %>%
  # Aggregate to individual-grade level  
  group_by(
    person_id,
    grade
  ) %>%
  summarise(
    gpa = mean(score, na.rm = TRUE)
  )
```

Since the main dataset is at individual level, we need to reshape the GPA to the level so the two datasets can be combined.

```{r}
gpa <-
  gpa %>%
  pivot_wider(
    id_cols = person_id,
    names_from = grade,
    names_prefix = "gpa_gr",
    values_from = gpa
  )
```

We can now combine the datasets.


```{r}
constructed <-
  constructed %>%
  left_join(
    gpa,
    by = "person_id"
  )
```

To confirm that the operation worked, we will look into the missing values created.

```{r}
constructed %>%
  select(starts_with("gpa")) %>%
  skim()
```

The number of missing values is exactly the same in both variables, indicating this is an issue with combining the two datasests. To confirm that both variables are always missing at the same time, we will do a cross tabulation.

```{r}
table(
  is.na(constructed$gpa_gr9), 
  is.na(constructed$gpa_gr10)
)
```

GPA information is always missing for the same people, which makes sense given the information we were provided about the data request: this information was only requested for defendants age 18 to 24. Weirdly enough, when we combine the datasets, the defendants age 18 to 24 are the only ones that we do not have GPA for:

```{r}
constructed %>%
  ggplot(
   aes(
     x = age,
     y = gpa_gr10
   )
  ) +
 geom_miss_point()
```

## Calculate number of prior arrests and re-arrests

We want to calculate the number of arrests prior to each case. This will be done in two parts: all arrests prior to the study period will be calculated first and then arrests during the study period will be added.


All of the individual’s arrests prior to the study period are contained in prior_arrest.csv. If someone is not included in prior_arrests.csv, they had zero arrests at the start of the study period. Some people really do get arrested multiple times on the same day. We will count each arrest separately, regardless of whether another arrest occurred on the same day. 

```{r}
prior_arrests <-
  prior_arrests %>%
  group_by(person_id) %>%
  summarise(
    preperiod_arrests = n()
  )
```

Some individuals were arrested multiple times during the study period and this should be accounted for in your prior arrest count.For example, if individual A was arrested 5 times prior to the study period and appears twice in the case file, their first arrest in the case file should have a prior arrest count of “5” and their second arrest should have a prior arrest count of “6”.

To calculate number of prior arrests in the study period, we will merge all the cases in the case dataset to all other cases from the same individual. We will then count the number of cases that happened before the arrest date. All the arrests that happen after the case’s arrest date and before the case’s disposition date are considered re-arrests, which we also want to calculate.


```{r}
study_period_arrests <-
  # Keep only person ID and date of arrest
  cases %>%
  select(
    person_id,
    date = arrest_date
  ) %>%
  # Merge to all arrests from the same person
  full_join(
    cases %>%
      select(
        person_id,
        caseid,
        arrest_date,
        dispos_date
      ),
    by = "person_id"
  ) %>%
  # For each combination of cases, create a dummy indicating whether
  # it is a prior arrest or a re-arrest
  mutate(
    prior_arrest = date < arrest_date,
    re_arrest = date > arrest_date & date < dispos_date
  )
```

For each case, we will now calculate the number of re-arrests and prior arrests during the study peiod by adding all the occurrences identified earlier.

```{r}
study_period_arrests <-
  study_period_arrests %>%
  group_by(caseid, person_id) %>%
  summarise(
    period_arrests = sum(prior_arrest, na.rm = TRUE),
    re_arrest = max(re_arrest) %>% as.logical()
  )
```

This information needs to be combined with the information on the pre-study period arrests to calculate the total number of prior arrests. 

```{r}
all_arrests <-
  study_period_arrests %>%
  # Combine both datasets
  left_join(
    prior_arrests,
    by = "person_id"
  ) %>%
  # Add pre-study and study period arrests
  rowwise() %>%
  mutate(
    prior_arrests_check = sum(
      period_arrests, preperiod_arrests, 
      na.rm = TRUE
    )
  )
```

To check that this calculation matches the original data received from the SA Office, we will also combine it to the raw checks data

```{r}
constructed <-
  constructed %>%
  left_join(
    all_arrests,
    by = c("person_id", "caseid")
  ) 
```

We will then include two statements to check that the values match. If they don't, the code will break.

```{r}
assert_that(
  all(
    constructed$prior_arrests.x == constructed$prior_arrests.y
  )
)

assert_that(
  all(
    constructed$re_arrest.x == constructed$re_arrest.y
  )
)
```

## Final dataset organization

Factors are usually sorted in alphabetical order. This means the base category for race will be "Asian". For comparison purposes, however, we want it to be "White". The next few lines of code change the order of the factors.

```{r}
constructed <-
  constructed %>%
  mutate(
    race = fct_relevel(
      race, 
      c("White", "Black", "Asian")
    )
  )
```

For our graphs and tables, we want treatment assignment values to be "treament" vs "control" instead of `TRUE` and `FALSE`, so we will rename the categories.

```{r}
constructed <-
  constructed %>%
  mutate(
    treat = case_when(
      treat == 1 ~ "Treatment",
      treat == 0 ~ "Control"
    )
  )
```

We will also remove from the dataset intermediate variables that we used to create indicators, but will not be used for analysis.

```{r}
constructed <-
  constructed %>%
  select(
    -c(
      st_address,
      city,
      bdate,
      -constains(".y")
    ) %>%
    prior_arrests = prior_arrests.x,
    re_arrest = re_arrest.x
  )
```

Finally, we will create variable labels to be shown in graphs and tables

```{r}
constructed <-
  constructed %>%
  var_labels(
    caseid = "Case ID",
    person_id = "Person ID",
    arrest_date = "Date of arrest",
    dispos_date = "Date of disposition",
    treat = "Treatment status",
    re_arrest = "Arrestee was rearrested",
    prior_arrests = "Number of arrestee's prior arrests",
    race = "Arrestee's race",
    gender = "Arrestee's gender",
    age = "Arrestee's age on arrest date"
  )
```

This is the dataset to be saved.

```{r}
write_rds(
  data,
  here(
    "data",
    "final",
    "analysis-data.rds"
  )
)
```
